Leaf traits drive differences in biomass partitioning among major plant functional types

### 2015-08-14 manuscript.Rnw
Fixed figure references; some minor text edits

We found that LMF differs predictably among PFTs, however there was considerable variation in all studied variables between species within PFTs (Fig. S4). It is thus pertinent to understand whether we can improve on a PFT-based classification by including other traits that affect biomass partitioning. To do this, we decomposed $M_F/A_S$ into LMA ($M_F/A_F$), and the ratio of leaf area to stem cross-sectional area ($A_F/A_S$) (Eq. \ref{eq:M2}).

@@

We found that decomposing LMF and LAR as shown in Eq. \ref{eq:M1} was very useful to study climate effects because the second term ($A_S/M_T$) absorbs nearly all of the size related variation in LMF (Tables \ref{table/varpartB}, Table S1 and Fig. S3) and is otherwise fairly conserved across PFTs (Fig. S3). As shown in Fig.~\ref{fig:figlmfspecies}b and Fig.~\ref{fig:figlmf}d, the term $M_F/A_S$ exhibits comparable differences between PFTs to LMF, but unlike LMF, $M_F/A_S$ is nearly independent of plant height (Tables \ref{table/varpartB} and Table S1). $M_F/A_S$ is thus a useful proxy for LMF that can be compared across species and sites.

@@

We found that PFTs differed in LMF, and that LMF was proportional to LMA. This conclusion was particularly strong across the three PFTs studied, but it also held across species within PFT. The implication is that the amount of leaf area supported per unit biomass (leaf area ratio, LAR) do not differ between PFTs. This result seems robust, as our data include individual plants spanning the entire size range of woody plants in natural forests (0.1-100m) (Fig.~\ref{fig:figlmf}a). Previous studies have demonstrated a large difference in LMF between angiosperms and gymnosperms \citep{poorter2012}, but have not further separated functional types into deciduous vs. evergreen angiosperms, or provided an explanation in terms of leaf traits. We show that LMA -- a central trait of the leaf economics spectrum (LES) \citep{wrightetal2004} -- explains these differences in a consistent manner.

A previous compilation of plant biomass partitioning \citep{poorter2012} (which included many non field-grown plants) did not find a correlation between LMF and LMA across 8 PFTs (including non-woody PFTs). One possible reason for this discrepancy, apart from differences in sample size and growing conditions, is that they did not correct for plant size in their comparison between LMF and LMA. We found that, as expected, plant size strongly influences LMF (Fig.~\ref{fig:figlmf}) and LAR (Fig. S3) (see also \citet{poorter2012}). It is thus necessary to correct for plant size when comparing biomass partitioning parameters, as has been noted many times \citep{mcconnaughay1999}. We used a semi-parametric approach to account for plant size, which has the advantage that it does not require an \emph{a priori} assumption on the functional relationship. This was useful because both LMF and LAR showed very non-linear patterns with plant height, even on a logarithmic scale.

@@

% The first sentence here feels a bit repetitive 
We found that, as for LMF, $M_F/A_S$ was also proportional to LMA across PFTs (Fig.~\ref{fig:figlmf}). Because LMF may be decomposed into the terms $M_F/A_S$ and $M_T/A_S$ (\ref{eq:M2}), this result suggests that $M_T/A_S$ does not vary between PFTs (see also Fig. S4). If this result holds, it means that aboveground biomass may be estimated across PFTs with an equation that does not differ between PFTs. For tropical forests, \citet{chave2005} showed that aboveground biomass can be well estimated as,

@@

where $F$ is a form factor, and $\rho$ wood density (kg m$^3$). After rearranging, we find that this equation predicts that $M_T/A_S$ is proportional to plant height. We found some support for this prediction (Fig. S4). It is timely to use the database made public by \citet{falster2015} (BAAD) to develop more general equations to estimate individual plant aboveground biomass across the globe.

We found that climate did not appreciably affect biomass partitioning between leaves and stems. This finding is in contrast with a recent study where LMF was found to increase with MAT across the globe \citep{reich2014}. One explanation for this difference is that our measurements were taken on individual plants, whereas in  \citet{reich2014} they were on whole stands. This suggests climatic effects on stand-level biomass partitioning occur primarily by altering the size-distribution and stand density (number of individuals per unit area), rather than partitioning within individual plants. With the caveats that there is still variation between species and studies unaccounted for (Fig. S5), and that coarse climate variables such as MAT and MAP may mask small-scale climate effects on biomass partitioning, our results strongly suggest the effects of PFT are stronger than any direct climate effect on biomass partitioning within individual plants (Tables \ref{table/varpartB} and Table S1).

Overall, our results indicate that it is possible to integrate a key leaf trait with whole-plant modelling of biomass partitioning. Indeed, nearly as much variation was explained across species when using LMA instead of PFT in a statistical model. This is surprising because LMA only captures one aspect of functional differentiation among PFT -- leaf morphology (but including correlated effects on leaf lifespan and photosynthetic capacity via the leaf economics spectrum \citep{wrightetal2004}). We thus show that in a PFT-based classification, LMA is a good first estimate of biomass partitioning, however, more variation can be explained by another trait ($A_F/A_S$), which varies appreciably between species. These results suggest promising avenues for parameterising and simplifying biomass partitioning routines in GVMs.

@@

\textbf{Table S1} Explained variance by plant functional type (PFT), plant height (H), mean annual temperature (MAT) and mean annual precipitation (MAP), for five whole-plant variables.\\
\textbf{Table S5} Number of observations used in the analysis of leaf area ratio (LAR).\\
\textbf{Fig. S1} Global coverage of the climate space by the dataset, labelled by vegetation type.\\
\textbf{Fig. S2} Leaf area ratio ($A_F/M_T$) by PFT.\\
\textbf{Fig. S3} Relationship between above-ground biomass and basal stem area.\\
\textbf{Fig. S4} Plant functional types diverge strongly in mass-based partitioning, but converge in area-based partitioning.\\
\textbf{Notes S1} Modelling of biomass partitioning in global vegetation models (GVMs)

>>>>

We found that LMF differs predictably among PFTs, however there was considerable variation in all studied variables between species within PFTs (Fig.\ref{fig:fighist}). It is thus pertinent to understand whether we can improve on a PFT-based classification by including other traits that affect biomass partitioning. To do this, we decomposed $M_F/A_S$ into LMA ($M_F/A_F$), and the ratio of leaf area to stem cross-sectional area ($A_F/A_S$) (Eq. \ref{eq:M2}).

@@

We found that decomposing LMF and LAR as shown in Eq. \ref{eq:M1} was very useful to study climate effects because the second term ($A_S/M_T$) absorbs nearly all of the size related variation in LMF (Tables \ref{table/varpartB}, Table S1 and Fig. S3) and is otherwise fairly conserved across PFTs (Fig.~\ref{fig:fighist}). As shown in Fig.~\ref{fig:figlmfspecies}b and Fig.~\ref{fig:figlmf}d, the term $M_F/A_S$ exhibits comparable differences between PFTs to LMF, but unlike LMF, $M_F/A_S$ is nearly independent of plant height (Tables \ref{table/varpartB} and Table S1). $M_F/A_S$ is thus a useful proxy for LMF that can be compared across species and sites.

@@

We found that PFTs differed in LMF, and that LMF was proportional to LMA. This conclusion was particularly strong across the three PFTs studied, but it also held across species within PFT. The implication is that the amount of leaf area supported per unit biomass (leaf area ratio, LAR) does not differ between PFTs. This result seems robust, as our data include individual plants spanning the entire size range of woody plants in natural forests (0.1-100m) (Fig.~\ref{fig:figlmf}a). Previous studies have demonstrated a large difference in LMF between angiosperms and gymnosperms \citep{poorter2012,poorter2015}, but have not further separated functional types into deciduous vs. evergreen angiosperms, or provided an explanation in terms of leaf traits. We show that LMA -- a central trait of the leaf economics spectrum (LES) \citep{wrightetal2004} -- explains these differences in a consistent manner.

A previous compilation of plant biomass partitioning \citep{poorter2012} (which included many non field-grown plants) did not find a correlation between LMF and LMA across 8 PFTs (including non-woody PFTs). One possible reason for this discrepancy, apart from differences in sample size and growing conditions, is that they did not correct for plant size in their comparison between LMF and LMA. We found that, as expected, plant size strongly influences LMF (Fig.~\ref{fig:figlmf}) and LAR (Fig. S2) (see also \citet{poorter2012}). It is thus necessary to correct for plant size when comparing biomass partitioning parameters, as has been noted many times \citep{mcconnaughay1999}. We used a semi-parametric approach to account for plant size, which has the advantage that it does not require an \emph{a priori} assumption on the functional relationship. This was useful because both LMF and LAR showed very non-linear patterns with plant height, even on a logarithmic scale.

@@

We found that patterns in $M_F/A_S$ across PFTs mirrored those of LMF (Fig.~\ref{fig:figlmf}). Because LMF may be decomposed into the terms $M_F/A_S$ and $M_T/A_S$ (\ref{eq:M2}), this result suggests that $M_T/A_S$ does not vary between PFTs (see also Fig. S3). If this result holds, it means that aboveground biomass may be estimated across PFTs with a single equation that does not differ between PFTs. For tropical forests, \citet{chave2005} showed that aboveground biomass can be well estimated as,

@@

where $F$ is a form factor, and $\rho$ wood density (kg m$^3$). After rearranging, we find that this equation predicts that $M_T/A_S$ is proportional to plant height. We found some support for this prediction (Fig. S3). It is timely to use database made public by \citet{falster2015} (BAAD) and \citet{poorter2015} to develop more general equations to estimate individual plant aboveground biomass across the globe.

We found that climate did not appreciably affect biomass partitioning between leaves and stems. This finding is in contrast with a recent study where LMF was found to increase with MAT across the globe \citep{reich2014}. One explanation for this difference is that our measurements were taken on individual plants, whereas in  \citet{reich2014} they were on whole stands. This suggests climatic effects on stand-level biomass partitioning occur primarily by altering the size-distribution and stand density (number of individuals per unit area), rather than partitioning within individual plants. With the caveats that there is still variation between species and studies unaccounted for (Fig.~\ref{fig:fighist}), and that coarse climate variables such as MAT and MAP may mask small-scale climate effects on biomass partitioning, our results strongly suggest the effects of PFT are stronger than any direct climate effect on biomass partitioning within individual plants (Tables \ref{table/varpartB} and Table S1).

Our results indicate that it is possible to integrate a key leaf trait with whole-plant modelling of biomass partitioning. Indeed, nearly as much variation was explained across species when using LMA instead of PFT in a statistical model. This is surprising because LMA only captures one aspect of functional differentiation among PFT -- leaf morphology (but including correlated effects on leaf lifespan and photosynthetic capacity via the leaf economics spectrum \citep{wrightetal2004}). We thus show that in a PFT-based classification, LMA is a good first estimate of biomass partitioning, however, more variation can be explained by another trait ($A_F/A_S$), which varies appreciably between species. These results suggest promising avenues for parameterising and simplifying biomass partitioning routines in GVMs.

@@

\textbf{Table S1} Explained variance by plant functional type (PFT), plant height (H), mean annual temperature (MAT) and mean annual precipitation (MAP), for five whole-plant variables.\\
\textbf{Table S5} Number of observations used in the analysis of leaf area ratio (LAR).\\
\textbf{Fig. S1} Global coverage of the climate space by the dataset, labelled by vegetation type.\\
\textbf{Fig. S2} Leaf area ratio ($A_F/M_T$) by PFT.\\
\textbf{Fig. S3} Relationship between above-ground biomass and basal stem area.\\
\textbf{Notes S1} Modelling of biomass partitioning in global vegetation models (GVMs)
